Proportions of Variants of Concern in NML Wastewater Data

### 2022-05-11 report/ms.tex
Example

\subsection{Motivating Example}

% Use the example from the presentation. It's good.
% This might be too "teachy" for a paper, though.

@@

In this formulation, $\rho$ is a latent vector that must be estimated.
Estimation is performed in two different ways.
A Bayesian model allows for inspection of posterior distributions and posterior covariance, as well as providing a basis for more complex analyses (see Conclusions and Future Work).
Note that the Bayesian model specification includes a Dirichlet prior to impose the positivity and sum-to-one constraints on $\rho$.
A constrained optimization is also implemented, which fits much faster and a bootstrap resampling scheme allows for estimation of confidence intervals.

>>>>

\subsection{Example}

To motivate our model choices, consider the following example.
Table \ref{toy_data} represents the format of our input data, showing the count and coverage of each mutation observed in the data.
The data also contain mutation aa:orf1a:S135R, which was \emph{not} observed in our data.
However, it is a mutation in one of the variants of concern and we were able to obtain the coverage at its position using the read depth for all positions (e.g., the command \texttt{samtools depth -aa}).

\begin{table}
\begin{tabular}{l|l|l}
Mutation & Count & Coverage\\\hline
aa:E:T91 & 1 & 2\\
aa:M:Q19E & 73 & 100\\
25000T & 387 & 1034\\
aa:orf1a:S135R & 0 & 3385
\end{tabular}
\caption{\label{toy_data}An example data set to motivate the model. Each row represents a single mutation, with the counts representing the number of times that mutation was observed and the coverage representing the read depth at that position.}
\end{table}

The variant matrix that we have chosen is shown in Table \ref{toy_varmat}. 
For this example we are only considering Omicron (B.1.1.529) and its subvariants BA.1 and BA.2.
The existence/absence of these mutations are based on the ``constellation'' lists provided by the PANGO team (\url{https://github.com/cov-lineages/constellations}).
This means that the mutation lists mostly consist of representative mutations; if a mutation is not on the list, it still may have been observed within that lineage but been removed because it was also present in other variants of concern.
As can be seen in Tab;e \ref{toy_varmat}, there are still overlaps in the mutation lists.

\begin{table}
\begin{tabular}{l|rrrrr}
& aa:E:T9I & aa:M:Q19E & 25000T & aa:orf1a:S135R\\\hline
B.1.1.529 & 1 & 1 & 0 & 0\\
BA.1 & 0 & 0 & 1 & 0\\
BA.2 & 0 & 0 & 1 & 1
\end{tabular}
\caption{\label{toy_varmat}An example of a variant matrix. The mutation list for the B.1.1.529 variant contains mutations aa:E:T9I and aa:M:Q19, hence the 1's in these columns, but does not contain the mutations 25000T or aa:orf1a:S135R.}
\end{table}

Consider the estimation of the proportion of B.1.1.529.
It contains the mutation aa:E:T9I, which was observed in 50\% of reads.
It also contains the mutation aa:M:Q19E, which was present in 73\% of reads.
Given this information alone, we may suppose that the proportion is halfway between these two proportions.
However, the 50\% is only based on 2 reads, while the 73\% is based on 100 reads.
An estimate of the proportion should be weighted higher when the read depth is higher.

Now lets consider BA.1 and BA.2.
These both contain the mutation 25000T, which is present in 37\% of the reads.
From this, we can guess that the proportion of BA.1 plus the proportion of BA.2 adds to approximately 37\%.
We also know that the mutation aa:orf1a:S135R was not observed in any reads and had a coverage of 3385, meaning that this mutation is very likely not present.
Since BA.2 contains this mutation, it is unlikely that BA.2 is present.
Since the proportions of BA.1 and BA.2 sum to 37\% and BA.2 is likely not present, BA.1 must have a proportion near 37\%.

This example highlights several features that our model includes.
First, we incorporate coverage information to give higher weight to mutations with higher coverage.
Where other methods remove mutations with low coverage, we simply give them very low weight in the estimation.
Second, we allow for shared mutations within mutation lists.
Third, we use information from mutations that were not present in the data.
A mutation that was not observed but has high coverage is evidence against the presence of a lineage, and this information can be leveraged to improve the estimate.

@@

In this formulation, $\rho$ is a latent vector that must be estimated.
Estimation is performed in two different ways.
A constrained optimization algorithm is implemented, which fits very quickly and a bootstrap resampling scheme allows for estimation of confidence intervals.
The constraints on this algorithm ensure that the proportions are positive and the sum of the proportions is between 0 and 1.
The

A Bayesian model allows for inspection of posterior distributions and posterior covariance, as well as providing a basis for more complex analyses (see Conclusions and Future Work).
Note that the Bayesian model specification includes a prior sampling scheme to impose the positivity and sum-to-one constraints on $\rho$.
In particular, if a proportion is sampled that causes the sum to be greater than one, it is rescaled so that the sum is exactly one.
This sampling ensures that the sum does not exceed 1, but is allowed to be less than 1. 
The prior is a beta distribution with parameters $\alpha=2$ and $\beta=8$, ensuring positivity of the proportions and low starting proportions for all lineages.
